Dependent Object Types

### 2012-09-15 doc/fool/article.tex
simplif abstract
intro rev

\begin{abstract}
We propose a new type-theoretic foundation of Scala and languages like
it: the Dependent Object Types calculus (DOT). DOT models Scala's
path-dependent types, abstract type members and its mixture of nominal
and structural typing through the use of refinement types.

@@

In this paper, we present and motivate the DOT calculus, our core
calculus for path-dependent types, both formally and informally. We
also discuss our work-in-progress to prove type-safety of the calculus.
\end{abstract}

@@

\section{Introduction}

This paper presents a proposal for a new type-theoretic foundation of
Scala and languages like it. The properties we are interested in
modeling are Scala's path-dependent types and abstract type members,
as well as its mixture of nominal and structural typing through the
use of refinement types. Compared to previous approaches~\cite{nuObj,FS},
we make no attempt to model inheritance or mixin composition. Indeed
we will argue that such concepts are better modeled in a different
setting.

The calculus does not precisely describe what's currently in Scala. It
is more normative than descriptive. The main point of deviation
concerns the difference between Scala's compound type formation using
{\bf with} and classical type intersection, as it is modeled in the
calculus. Scala, and the previous calculi attempting to model it,
conflates the concepts of compound types (which inherit the members of
several parent types) and mixin composition (which build classes from
other classes and traits). At first glance, this offers an economy of
concepts. However, it is problematic because mixin composition and
intersection types have quite different properties. In the case of
several inherited members with the same name, mixin composition has to
pick one which overrides the others. It uses for that the concept of
linearization of a trait hierarchy. Typically, given two independent
traits $T_1$ and $T_2$ with a common method $m$, the mixin composition
\code@$T_1$ with $T_2$@ would pick the $m$ in $T_2$, whereas the member in
$T_1$ would be available via a super-call. All this makes sense from
an implementation standpoint. From a typing standpoint it is more
awkward, because it breaks commutativity and with it several
monotonicity properties.

@@

This is problematic because glbs and lubs play a central role in
Scala's type inference. For example, in to infer the type \code{if} expression such as
\begin{lstlisting}
  if (cond) ((a: A) => c: C) else ((b: B) => d: D)
\end{lstlisting}
type inference tries to compute the glb of \code{A} and \code{B} and
the lub of \code{C} and \code{D} The absence of universal glbs and
lubs makes type inference more brittle and more unpredictable.

%The primary contributions of this paper are:
%\begin{itemize}
%\item Development of DOT -- a core calculus which models
%  path-dependent types. We formally specify a small-step operational
%  semantics and a static typing assignment in
%  section~\ref{dot-calculus} and show some examples in
%  section~\ref{dot-examples}.
%\item Counterexamples to the standard theorem of preservation --
%  section~\ref{dot-preservation}.
%\item A sketch of a type-safety proof based on logical relations --
%  section~\ref{dot-type-safety}.
%\end{itemize}

We propose DOT as a core calculus for path-dependent types. We first
motivate the calculus informally in section~\ref{dot-motiv}. We
present the calculus formally in section \ref{dot-calculus} and
through examples in section~\ref{dot-examples}. Though we show that
the calculus does not satisfy the standard theorem of preservation in
section~\ref{dot-preservation}, we contribute a proof sketch of type
safety using logical relations in section~\ref{dot-type-safety}. In
section~\ref{discussion}, we discuss choices and variants of the
calculus, as well as related work, and conclude in
section~\ref{conclusion}.

\section{Motivation}\label{dot-motiv}
%TODO

@@

>>>>

\begin{abstract}
We propose a new type-theoretic foundation of Scala and languages like
it: the Dependent Object Types (DOT) calculus. DOT models Scala's
path-dependent types, abstract type members and its mixture of nominal
and structural typing through the use of refinement types.

@@

In this paper, we present the DOT calculus, both formally and
informally. We also discuss our work-in-progress to prove type-safety
of the calculus.
\end{abstract}

@@

\section{Introduction}
A scalable programming language is one in which the same concepts can
describe small as well as large parts. Towards this goal, Scala
unifies concepts from object and module systems. An essential
ingredient of this unification is objects with type members. Given a
stable path to an object, its type members can be accessed as types,
called path-dependent types.

This paper presents Dependent Object Types (DOT), a small object
calculus with path-dependent types. In addition to path-dependent
types, types in DOT are built from refinements, intersections and
unions. A refinement extends a type by (re-)declaring members, which
can be types, values or methods.

We propose DOT as a new type-theoretic foundation of Scala and
languages like it. The properties we are interested in modeling are
Scala's path-dependent types and abstract type members, as well as its
mixture of nominal and structural typing. Compared to previous
approaches~\cite{nuObj,FS}, we make no attempt to model inheritance or
mixin composition. Indeed we will argue that such concepts are better
modeled in a different setting.

The DOT calculus does not precisely describe what's currently in Scala. It
is more normative than descriptive. The main point of deviation
concerns the difference between Scala's compound type formation using
{\bf with} and classical type intersection, as it is modeled in the
calculus. Scala, and the previous calculi attempting to model it,
conflates the concepts of compound types (which inherit the members of
several parent types) and mixin composition (which build classes from
other classes and traits). At first glance, this offers an economy of
concepts. However, it is problematic because mixin composition and
intersection types have quite different properties. In the case of
several inherited members with the same name, mixin composition has to
pick one which overrides the others. It uses for that the concept of
linearization of a trait hierarchy. Typically, given two independent
traits $T_1$ and $T_2$ with a common method $m$, the mixin composition
\code@$T_1$ with $T_2$@ would pick the $m$ in $T_2$, whereas the member in
$T_1$ would be available via a super-call. All this makes sense from
an implementation standpoint. From a typing standpoint it is more
awkward, because it breaks commutativity and with it several
monotonicity properties.

@@

This is problematic because glbs and lubs play a central role in
Scala's type inference. For example, in order to infer the type of an
\code{if} expression such as
\begin{lstlisting}
  if (cond) ((a: A) => c: C) else ((b: B) => d: D)
\end{lstlisting}
type inference tries to compute the glb of \code{A} and \code{B} and
the lub of \code{C} and \code{D}. The absence of universal glbs and
lubs makes type inference more brittle and more unpredictable.

Compared to Scala, DOT also simplifies type members. In DOT, a type
member is declared by its lower and upper bounds. Furthermore, unlike
Scala, there is no notion of uninstiated type member. Here is an
example inspired by~\cite{vt}:
\begin{lstlisting}
trait Food
trait Animal {
  type Meal <: Food
  def eat(meal: Meal) {}
}
\end{lstlisting}
In Scala, \code{Meal} is an uninstiated type member of \code{Animal}:
it is not possible to instantiate an \code{Animal} without providing a
concrete binding for \code{Meal}. In DOT, \code{Meal} is a type member
with a lower bound of $\Bot$ and an upper bound of \code{Food}. It is
possible to instantiate an \code{Animal} without further specifying
\code{Meal}, though it would be not possible to feed it. Now,
we define \code{Cow} by refining \code{Animal}.
\begin{lstlisting}
trait Grass extends Food
trait Cow extends Animal {
  type Meal = Grass
}
\end{lstlisting}
In Scala, the type alias \code{type Meal = Grass} defines a concrete
binding for \code{Meal}. Now, we can instantiate \code{Cow}s and feed
them \code{Grass}. In DOT, such a type alias is declared by giving it
identical lower and upper bounds.

DOT has a notion of concrete vs abstract type members, but it used to
distinguish which types members are instantiable as opposed to
instantiated. In the example above, all types introduced by
\code{trait} would be concrete type members in DOT with lower bounds
of $\Bot$ and upper bounds describing the \code{extends} clause:
$\Top$ for \code{Food}, a refinement of $\Top$ with type member
\code{Meal} and method member \code{eat} for \code{Animal},
\code{Food} for \code{Grass} and a refinement of \code{Animal} with
type member \code{Meal} for \code{Cow}. Concrete type members
typically have a lower bound of $\Bot$ so that they are purely
nominal: e.g. one needs to directly or indirectly instantiate a
\code{Cow} to have an object of type \code{Cow}.

We propose DOT as a core calculus for path-dependent types. We present
the calculus formally in section \ref{dot-calculus} and through
examples in section~\ref{dot-examples}. Though we show that the
calculus does not satisfy the standard theorem of preservation in
section~\ref{dot-preservation}, we contribute a proof sketch of type
safety using logical relations in section~\ref{dot-type-safety}. In
section~\ref{discussion}, we discuss choices and variants of the
calculus, as well as related work, and conclude in
section~\ref{conclusion}.

@@

\subsection{Scala to DOT}\label{dot-examples-scala-sugar}
// TODO
